---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# streetlamp

`streetlamp` is an R package for estimating the effects of police stop and
search activity on recorded crime in England and Wales, published by Black
Thrive Global. It builds an area-by-month panel of recorded crime from the
data.police.uk archive, attaches stop and search intensity as a treatment
variable, and provides a panel-econometrics toolkit for evaluating policing
interventions: event studies, difference-in-differences with staggered
adoption, synthetic control, spatial spillover and displacement models, and
stop-crime elasticities by offence type.

`streetlamp` is about what stops achieve, not who is stopped. It never
computes an ethnic disparity measure and reads stop and search files only to
count stops by area and month.

## Installation

The package is under development and not yet on CRAN.

``` r
# install.packages("pak")
pak::pak("black-thrive-global/streetlamp")
```

## Status

Milestones M0 (skeleton) and M1 (ingestion and versioning) are complete:
package infrastructure, continuous integration, the bundled reference tables,
selective download from the data.police.uk archive, one-version-per-file
selection with the alternatives recorded, and readers for crime, outcomes and
stop counts that carry a panel contract. Panel construction and the coverage
audit (M2), treatment definitions and estimators (M3 and M4) and reporting
(M5) follow.

## Reading the archive

The data.police.uk archive publishes one zip per month, each 1 to 2.6 GB.
`streetlamp` reads a zip's table of contents with an HTTP byte-range request
and fetches only the force-month files you ask for, so a three-year pull for
two forces is tens of megabytes. The package ships two miniature archive
snapshots (real records for a subset of West Yorkshire and Dyfed-Powys
neighbourhoods, May to July 2026), so everything below runs offline.

```{r}
library(streetlamp)

cache <- tempfile("streetlamp-cache-")
zips <- list.files(
  system.file("extdata", "archive", package = "streetlamp"),
  pattern = "zip$", full.names = TRUE
)
for (z in zips) lamp_archive_register(z, dir = cache)
snap <- lamp_archive_snapshot(cache)
snap

# The same force-month can appear in several snapshots with different content
versions <- lamp_list_versions(snap)
versions[versions$differs, c("force_id", "month", "file_type", "n_versions")]
lamp_version_diff(snap, "west-yorkshire", "2026-05", "outcomes")

# Readers return tibbles with a fixed schema and a contract
crime <- lamp_read_crime(snap, forces = "dyfed-powys")
table(crime$crime_type)
lamp_contract(crime)$coverage

stops <- lamp_read_stop_counts(snap)
stops
```

With network access, `lamp_archive_download()` fetches force-month files from
the live archive into the cache:

``` r
snap <- lamp_archive_download(
  archives = "latest",
  months = c("2026-05", "2026-06", "2026-07"),
  forces = c("west-yorkshire", "dyfed-powys")
)
```

## Bundled reference tables

Everything in the package runs without network access at check time. The
reference tables shipped with it are verified against their sources, as
recorded in `inst/NOTES/data_sources.md`.

```{r}
library(streetlamp)

# The fourteen data.police.uk categories with package groupings
lamp_crime_types()

# Dated national shocks for use as default intervention definitions
shocks <- lamp_shocks()
shocks[shocks$kind == "stop_search_policy", c("name", "start", "end", "scope")]

# Bank holidays in England and Wales, for seasonality controls
range(lamp_bank_holidays()$date)

# ONS LSOA code lists and the 2011 to 2021 lookup
length(lamp_lsoa_codes("lsoa11"))
length(lamp_lsoa_codes("lsoa21"))
table(lamp_lsoa_lookup()$change)
```

## Design principles

* Every panel carries a **panel contract** recording provenance, coverage,
  LSOA vintage and treatment definition. Estimators read it and refuse or
  warn when their assumptions are violated.
* "Force did not submit" is never a zero.
* Anti-social behaviour is not a crime: it is kept in a separate series and
  excluded from crime totals by default.
* Every estimator names its identifying assumption in its output and ships
  with a placebo or pre-trend diagnostic.

## Data attribution

Contains public sector information licensed under the Open Government Licence
v3.0: recorded crime, outcomes and stop and search data from
[data.police.uk](https://data.police.uk/), geography from the
[Office for National Statistics](https://geoportal.statistics.gov.uk/), and
bank holidays from [GOV.UK](https://www.gov.uk/bank-holidays).

## Licence

MIT, copyright Black Thrive Global.
