---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# streetlamp

`streetlamp` is an R package for estimating the effects of police stop and
search activity on recorded crime in England and Wales, published by Black
Thrive Global. It builds an area-by-month panel of recorded crime from the
data.police.uk archive, attaches stop and search intensity as a treatment
variable, and provides a panel-econometrics toolkit for evaluating policing
interventions: event studies, difference-in-differences with staggered
adoption, synthetic control, spatial spillover and displacement models, and
stop-crime elasticities by offence type.

`streetlamp` is about what stops achieve, not who is stopped. It never
computes an ethnic disparity measure and reads stop and search files only to
count stops by area and month.

## Installation

The package is under development and not yet on CRAN.

``` r
# install.packages("pak")
pak::pak("BlackThrive/streetlamp")
```

## Status

The 0.2.0 development series adds count-mean elasticities, calendar-aware
lags, heterogeneous simulation designs, covariance-aware effect summaries,
design audits and complete robustness grids. Source and continuous
integration are available on GitHub. Historical 0.1.0 validation is retained;
new research evidence is in `inst/validation/research-upgrade`. A causal
empirical case still needs independently documented intervention assignment.
This development version has not been submitted to CRAN.
Cross-platform release checks for the expanded development version are pending.

## What it estimates

| Question | Function |
|---|---|
| Did crime fall after a dated intervention? | `lamp_event_study()`, `lamp_twfe()` |
| Areas adopted at different dates | `lamp_did_staggered()` |
| Did crime move next door instead? | `lamp_spillover()`, `lamp_displacement_quotient()` |
| One treated force, many comparison areas | `lamp_synth()` |
| How does crime respond to search intensity? | `lamp_elasticity()` |
| How does searching follow crime? | `lamp_allocation()` |
| How many crimes per thousand searches? | `lamp_crimes_prevented()` |
| Is the design supported by the observed data? | `lamp_design_audit()` |
| Is the result sensitive to specifications or individual areas? | `lamp_elasticity_robustness()` |
| What is the uncertainty of an average or sum of effects? | `lamp_effect_summary()` |
| How would an assumed differential trend change an event estimate? | `lamp_trend_sensitivity()` |
| Is the estimate worth anything? | `lamp_pretrends()`, `lamp_placebo()` |

`lamp_elasticity(method = "fe")` now defaults to Poisson pseudo-likelihood
for the conditional count mean. Use `family = "ols_log"` for the former
response scale, which CCE methods retain. The covariate is `log(1 + stops)`:
its slope is not automatically a conventional stops elasticity. At stop
intensity S, a count-mean slope becomes `slope * S / (1 + S)`. Observational
associations do not identify the effect of deciding to search more.

## Quick start on the bundled panel

The package ships a ready-made panel: every 2021 LSOA in West Yorkshire and
Dyfed-Powys over August 2024 to July 2026, with crime by type, anti-social
behaviour, stops, population and deprivation deciles, and a contract that
records where every number came from.

```{r}
library(streetlamp)

panel <- lamp_sample_panel()
panel
lamp_contract(panel)

# Coverage: which force-months exist, and where stop files are missing while
# crime files are present
cov <- lamp_coverage(panel)
table(cov$file_type, cov$status)
lamp_coverage_compare(cov, c("2024-08", "2025-07"), c("2025-08", "2026-07"))

# Spatial structure for the spillover estimators
adj <- lamp_adjacency(lamp_sample_boundaries())
adj
```

The plot methods draw force-month totals with missing months shaded
(`plot(panel)`) and the coverage grid as a heat map (`plot(cov)`).

## Estimating, and checking the estimate

```{r estimate}
# a dated intervention on a simulated panel whose true effect is known
sim <- lamp_simulate(n_areas = 60, n_months = 24, design = "event", effect = -0.25, seed = 1)
truth <- attr(sim, "truth")
tr <- lamp_treatment(sim, "event", date = truth$event_date, scope = truth$treated_areas)

fit <- lamp_twfe(sim, "crime_total", tr)
fit

# the true effect on the crime total, for comparison
truth$effect_crime_total
```

Every estimate prints the assumption it rests on and the force-months it had
to drop. The diagnostics are the point of the package as much as the
estimators are:

```{r diagnostics}
es <- lamp_event_study(sim, "crime_total", tr, window = c(-6, 6))

# how large a pre-trend would this test have missed, and what would it cost?
lamp_pretrends(es)$power

# does the specification produce effects where none exist?
lamp_placebo(fit, type = "space", n = 50, seed = 1)$p_value
```

## Reading the results honestly

Police send officers where crime has risen, so searching responds to crime as
well as possibly reducing it. `lamp_allocation()` measures that channel, and
nothing else in the package should be read without it.

```{r allocation}
lamp_allocation(sim, crime_lags = 1:3)$diagnostics$interpretation
```

`inst/NOTES/methods.md` sets out the equations, the assumption behind each
estimator, and the limits that apply to all of them: recording practice,
anonymised locations, and the fact that recorded crime is what the police
wrote down.

## Reading the archive

The data.police.uk archive publishes one zip per month, each 1 to 2.6 GB.
`streetlamp` reads a zip's table of contents with an HTTP byte-range request
and fetches only the force-month files you ask for, so a three-year pull for
two forces is tens of megabytes. The package ships two miniature archive
snapshots (real records for a subset of West Yorkshire and Dyfed-Powys
neighbourhoods, May to July 2026), so everything below runs offline.

```{r}
cache <- tempfile("streetlamp-cache-")
zips <- list.files(
  system.file("extdata", "archive", package = "streetlamp"),
  pattern = "zip$", full.names = TRUE
)
for (z in zips) lamp_archive_register(z, dir = cache)
snap <- lamp_archive_snapshot(cache)
snap

# The same force-month can appear in several snapshots with different content
versions <- lamp_list_versions(snap)
versions[versions$differs, c("force_id", "month", "file_type", "n_versions")]
lamp_version_diff(snap, "west-yorkshire", "2026-05", "outcomes")

# Readers return tibbles with a fixed schema and a contract
crime <- lamp_read_crime(snap, forces = "dyfed-powys")
table(crime$crime_type)
lamp_contract(crime)$coverage

stops <- lamp_read_stop_counts(snap)
stops
```

With network access, `lamp_archive_download()` fetches force-month files from
the live archive into the cache, `lamp_boundaries()` and `lamp_population()`
fetch ONS boundaries and Census 2021 population once, and `lamp_panel()`
builds the balanced panel:

``` r
snap <- lamp_archive_download(
  archives = "latest",
  months = c("2026-05", "2026-06", "2026-07"),
  forces = c("west-yorkshire", "dyfed-powys")
)
crime <- lamp_read_crime(snap)
stops <- lamp_read_stop_counts(snap, area = "lsoa21", boundaries = lamp_boundaries("lsoa21"))
panel <- lamp_panel(crime, stops = stops, population = lamp_population("lsoa21"))
```

## Bundled reference tables

Everything in the package runs without network access at check time. The
reference tables shipped with it are verified against their sources, as
recorded in `inst/NOTES/data_sources.md`.

```{r}
library(streetlamp)

# The fourteen data.police.uk categories with package groupings
lamp_crime_types()

# Dated national shocks for use as default intervention definitions
shocks <- lamp_shocks()
shocks[shocks$kind == "stop_search_policy", c("name", "start", "end", "scope")]

# Bank holidays in England and Wales, for seasonality controls
range(lamp_bank_holidays()$date)

# ONS LSOA code lists and the 2011 to 2021 lookup
length(lamp_lsoa_codes("lsoa11"))
length(lamp_lsoa_codes("lsoa21"))
table(lamp_lsoa_lookup()$change)
```

## Design principles

* Every panel carries a **panel contract** recording provenance, coverage,
  LSOA vintage and treatment definition. Estimators read it and refuse or
  warn when their assumptions are violated.
* "Force did not submit" is never a zero.
* Anti-social behaviour is not a crime: it is kept in a separate series and
  excluded from crime totals by default.
* Every estimator names its identifying assumption in its output and ships
  with a placebo or pre-trend diagnostic.

## Data attribution

Contains public sector information licensed under the Open Government Licence
v3.0: recorded crime, outcomes and stop and search data from
[data.police.uk](https://data.police.uk/), geography from the
[Office for National Statistics](https://geoportal.statistics.gov.uk/), and
bank holidays from [GOV.UK](https://www.gov.uk/bank-holidays).

## Licence

MIT, copyright Black Thrive Global.
