---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# streetlamp

`streetlamp` is an R package for estimating the effects of police stop and
search activity on recorded crime in England and Wales, published by Black
Thrive Global. It builds an area-by-month panel of recorded crime from the
data.police.uk archive, attaches stop and search intensity as a treatment
variable, and provides a panel-econometrics toolkit for evaluating policing
interventions: event studies, difference-in-differences with staggered
adoption, synthetic control, spatial spillover and displacement models, and
stop-crime elasticities by offence type.

`streetlamp` is about what stops achieve, not who is stopped. It never
computes an ethnic disparity measure and reads stop and search files only to
count stops by area and month.

## Installation

The package is under development and not yet on CRAN.

``` r
# install.packages("pak")
pak::pak("black-thrive-global/streetlamp")
```

## Status

Milestone M0 (skeleton) is complete: package infrastructure, continuous
integration, and the bundled reference tables below. Ingestion of the
data.police.uk archive (M1), panel construction and coverage audit (M2),
treatment definitions and estimators (M3 and M4) and reporting (M5) follow.

## Bundled reference tables

Everything in the package runs without network access at check time. The
reference tables shipped with it are verified against their sources, as
recorded in `inst/NOTES/data_sources.md`.

```{r}
library(streetlamp)

# The fourteen data.police.uk categories with package groupings
lamp_crime_types()

# Dated national shocks for use as default intervention definitions
shocks <- lamp_shocks()
shocks[shocks$kind == "stop_search_policy", c("name", "start", "end", "scope")]

# Bank holidays in England and Wales, for seasonality controls
range(lamp_bank_holidays()$date)

# ONS LSOA code lists and the 2011 to 2021 lookup
length(lamp_lsoa_codes("lsoa11"))
length(lamp_lsoa_codes("lsoa21"))
table(lamp_lsoa_lookup()$change)
```

## Design principles

* Every panel carries a **panel contract** recording provenance, coverage,
  LSOA vintage and treatment definition. Estimators read it and refuse or
  warn when their assumptions are violated.
* "Force did not submit" is never a zero.
* Anti-social behaviour is not a crime: it is kept in a separate series and
  excluded from crime totals by default.
* Every estimator names its identifying assumption in its output and ships
  with a placebo or pre-trend diagnostic.

## Data attribution

Contains public sector information licensed under the Open Government Licence
v3.0: recorded crime, outcomes and stop and search data from
[data.police.uk](https://data.police.uk/), geography from the
[Office for National Statistics](https://geoportal.statistics.gov.uk/), and
bank holidays from [GOV.UK](https://www.gov.uk/bank-holidays).

## Licence

MIT, copyright Black Thrive Global.
